Notes on Validated Model Counting \\ Version of \today

### 2022-11-09 doc/mc-notes.tex
Clarify requirements for final status

The CRAT format draws its inspiration from the LRAT format for Boolean formulas and the
QRAT format for
quantified Boolean formulas (QBF).  The following are its key properties:
\begin{itemize}
  \item
  In addition to explicit clause additions and deletions, the proof contains declarations of $\pand$ and $\por$ operations.
\begin{itemize}
\item 
  These declarations implicitly add extension variables and their defining clauses to the proof.
  This is the only means for generating extension variables or adding blocked clauses to the proof.
\item The checker tracks the dependency set for every input and
  extension variable.  When an extension variable is introduced based
  on the $\pand$ operation, the dependency sets of its arguments must
  be disjoint.  The resulting extension variable has a dependency set equal to the union of those of its arguments.
\item Declaring a $\por$ operation requires a sequence of clauses
  providing a RUP proof that the arguments are mutually exclusive.  This sequence can either be provided explicitly or inferred by the proof checker.
\item Boolean complement is provided implicitly by allowing the
  arguments of the $\por$ and $\pand$ operations to be literals and not just
  variables.
\end{itemize}
\item
  Clauses can also be added when they satisfy the RUP
  property, with respect to a sequence of existing clauses.  This sequence can
  be either supplied explicitly or inferred by the proof checker.
\item
  Deleting clauses requires proving that the resulting formula is not weaker.
  \begin{itemize}
    \item For an input clause or a clause declared by RUP addition, its
      deletion must be accompanied by a sequence of remaining clauses
      providing a RUP proof of the clause.  This sequence can either
      be provided explicitly or inferred by the proof checker.
    \item The clauses defining a $\pand$ or $\por$ operation are
      implicitly deleted by deleting the operation.  This can only be
      done when the only undeleted clauses containing
      references to the associated extension variable are those
      implicitly defined when the operation was declared.
      This implies that there can be no undeleted operations having
      the operation result as an argument.
  \end{itemize}
\end{itemize}

@@

\begin{table}
  \caption{CRAT Step Types.  $C$: clause identifier, $L$: literal, $V$: variable}
  \label{tab:crat:syntax}
\centering{
  \begin{tabular}{lllll}
    \multicolumn{4}{c}{Rule} & \multicolumn{1}{c}{Description} \\
    \midrule
    \makebox[5mm][l]{$C$} & \makebox[10mm][l]{\tt i}  & \makebox[15mm][l]{$L^{*}$ {\tt 0}} & \makebox[15mm][l]{}  & \makebox[20mm][l]{Input clause} \\
    $C$ & {\tt a} & $L^{*}$ {\tt 0} & $C^{+}$  {\tt 0} & Add RUP clause.  Explicit hint \\
    $C$ & {\tt a} & $L^{*}$ {\tt 0} &  {\tt * 0} & Add RUP clause.  Inferred hint \\
     & {\tt dc} & $C$             & $C^{+}$  {\tt 0} & Delete RUP clause.  Explicit hint \\
     & {\tt dc} & $C$             & {\tt * 0} & Delete RUP clause.  Inferred hint \\
    \midrule
    $C$    & {\tt p} & $V \; L^{2+}$ {\tt 0}    &                  & Declare $\pand$ operation \\
    $C$    & {\tt s} & $V \; L \; L$    & $C^{+}$ {\tt 0}  & Declare $\por$ operation.  Explicit hint \\ 
    $C$    & {\tt s} & $V \; L \; L$    & {\tt * 0}  & Declare $\por$ operation.  Inferred hint \\ 
%%        & {\tt ite} & $V \; L \; L\; L$    &                  & Declare $\pite$ operation \\
     & {\tt do} & $V$             &  & Delete operation. \\

@@

Table~\ref{tab:crat:syntax} shows the declarations that can occur in a CRAT file.
As with other clausal proof formats, a variable is
represented by a positive integer $v$, with the first ones being input
variables and successive ones being extension variables.  Literal $l$
is represented by a signed integer, with $-v$ being the complement of
variable $v$.  Each clause is indicated by a positive integer
identifier $C$, with the first ones being the IDs of the input clauses and successive
ones being the IDs of added clauses.  Clause identifiers must be totally ordered,
such that clause $C$ can only reference clauses $C'$ such that $C' < C$.
Clause identifiers need not be consecutive.

The first set of proof rules are similar to those in other clausal
proofs.  Our syntax optionally allows input clauses to be listed with
a rule of type {\tt i}.  Clauses can be added via RUP addition
(command {\tt a}), with the sequence of antecedent clauses (the
``hint'') either provided explicitly or to be inferred (indicated by the character `{\tt *}') by the proof
checker.  Similarly for clause deletion (command {\tt dc}).

@@

A CRAT proof follows the same general form as a QRAT dual
proof~\cite{bryant:cade:2021}---one that ensures that each clause
addition and each clause deletion preserves equivalence.  With CRAT,
however, clauses are defined both explicitly and implicitly.  Starting
with the set of input clauses, the proof consists of a sequence of steps that
both add and delete clauses.  Each addition must be truth preserving
and each deletion must be falsehood preserving.  At the end, all input
clauses must have been deleted, and among the remaining clauses there
must be only a single unit clause consisting of some variable or its
complement.  Except for trivial cases, the final literal will be an
extension variable or its complement.  That literal will indicate the
root of the POG as the (possibly negated) output of a POG
operation.  Working from that root backward, the POG can be
extracted from the CRAT file.

>>>>

The CRAT format draws its inspiration from the LRAT format for Boolean formulas and the
QRAT format for
quantified Boolean formulas (QBF).  The following are its key properties:
\begin{itemize}
  \item
  In addition to explicit clause additions and deletions, the proof contains declarations of $\pand$ and $\por$ operations.
\begin{itemize}
\item 
  These declarations implicitly add extension variables and their defining clauses to the proof.
  This is the only means for generating extension variables or adding blocked clauses to the proof.
\item The checker tracks the dependency set for every input and
  extension variable.  When an extension variable is introduced based
  on the $\pand$ operation, the dependency sets of its arguments must
  be disjoint.  The resulting extension variable has a dependency set equal to the union of those of its arguments.
\item Declaring a $\por$ operation requires a sequence of clauses
  providing a RUP proof that the arguments are mutually exclusive.  This sequence can either be provided explicitly or inferred by the proof checker.
\item Boolean complement is provided implicitly by allowing the
  arguments of the $\por$ and $\pand$ operations to be literals and not just
  variables.
\end{itemize}
\item
  Clauses can also be added when they satisfy the RUP
  property, with respect to a sequence of existing clauses.  This sequence can
  be either supplied explicitly or inferred by the proof checker.
\item
  Deleting clauses requires proving that the resulting formula is not weaker.
  That is, it should not allow any additional satisfying assignment.
  \begin{itemize}
    \item For an input clause or a clause declared by RUP addition, its
      deletion must be accompanied by a sequence of remaining clauses
      providing a RUP proof of the clause.  This sequence can either
      be provided explicitly or inferred by the proof checker.
    \item The clauses defining a $\pand$ or $\por$ operation are
      implicitly deleted by deleting the operation.  This can only be
      done when the only undeleted clauses containing
      references to the associated extension variable are those
      implicitly defined when the operation was declared.
      This implies that there can be no undeleted operations having
      the operation result as an argument.
  \end{itemize}
\end{itemize}

@@

\begin{table}
  \caption{CRAT Step Types.  $C$: clause identifier, $L$: literal, $V$: variable}
  \label{tab:crat:syntax}
\centering{
  \begin{tabular}{lllll}
    \multicolumn{4}{c}{Rule} & \multicolumn{1}{c}{Description} \\
    \midrule
    \makebox[5mm][l]{$C$} & \makebox[10mm][l]{\tt a}  & \makebox[15mm][l]{$L^{*}$ {\tt 0}} & \makebox[15mm][l]{$C^{+}$ {\tt 0}}  & \makebox[20mm][l]{Add RUP clause.  Explicit hint} \\
    $C$ & {\tt a} & $L^{*}$ {\tt 0} &  {\tt * 0} & Add RUP clause.  Inferred hint \\
     & {\tt dc} & $C$             & $C^{+}$  {\tt 0} & Delete RUP clause.  Explicit hint \\
     & {\tt dc} & $C$             & {\tt * 0} & Delete RUP clause.  Inferred hint \\
    \midrule
    $C$    & {\tt p} & $V \; L^{2+}$ {\tt 0}    &                  & Declare $\pand$ operation \\
    $C$    & {\tt s} & $V \; L \; L$    & $C^{+}$ {\tt 0}  & Declare $\por$ operation.  Explicit hint \\ 
    $C$    & {\tt s} & $V \; L \; L$    & {\tt * 0}  & Declare $\por$ operation.  Inferred hint \\ 
%%        & {\tt ite} & $V \; L \; L\; L$    &                  & Declare $\pite$ operation \\
     & {\tt do} & $V$             &  & Delete operation. \\

@@

Table~\ref{tab:crat:syntax} shows the declarations that can occur in a CRAT file.
The checker is provided with the input formula as a separate file.
As with other clausal proof formats, a variable is
represented by a positive integer $v$, with the first ones being input
variables and successive ones being extension variables.  Literal $l$
is represented by a signed integer, with $-v$ being the complement of
variable $v$.  Each clause is indicated by a positive integer
identifier $C$, with the first ones being the IDs of the input clauses and successive
ones being the IDs of added clauses.  Clause identifiers must be totally ordered,
such that clause $C$ can only reference clauses $C'$ such that $C' < C$.
Clause identifiers need not be consecutive.

The first set of proof rules are similar to those in other clausal
proofs.
Clauses can be added via RUP addition
(command {\tt a}), with the sequence of antecedent clauses (the
``hint'') either provided explicitly or to be inferred (indicated by the character `{\tt *}') by the proof
checker.  Similarly for clause deletion (command {\tt dc}).

@@

A CRAT proof follows the same general form as a QRAT dual
proof~\cite{bryant:cade:2021}---one that ensures that each clause
addition and each clause deletion preserves equivalence.  With CRAT,
however, clauses are defined both explicitly and implicitly.  Starting
with the set of input clauses, the proof consists of a sequence of steps that
both add and delete clauses.  Each addition must be truth preserving
and each deletion must be falsehood preserving.  At the end, the following conditions must hold:
\begin{itemize}
\item All of the input clauses should have been deleted.
\item There should be exactly one clause left that was added via RUP addition, and this should be a unit clause.
\end{itemize}
The remaining
unit clause consists of some variable or its
complement.  Except for trivial cases, the final literal will be an
extension variable or its complement.  That literal will indicate the
root of the POG as the (possibly negated) output of a POG
operation.  Working from that root backward, the POG can be
extracted from the CRAT file.
